---
title: | 
  | Demo Job intterviw at FE.
  | Data Scientist
author: "Youssouf Souare"
date:  "August 28, 2017"
output:
  html_document: 
    number_sections: true
  pdf_document: 
    number_sections: true
  word_document:
    number_sections: true
    dev: png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, dpi = 300)

```

```{r  echo= FALSE}
##########################
library(tidyverse) 
library(lubridate) # utils for data manupulation
library(pander) # for prettier tables
library(scales) # for making prettier axes in plots
library(plotly)  # utils for interaction ploting 
library(devtools)
library(maps)
library(corrgram) # to investigate the correlation between variables
library(corrplot) # to plot correlation data
library(knitr) # utils for html table styling
library(kableExtra) # utils for html table styling
library(caTools)
library(modelr)
library(GGally)
library(viridis)
library(factoextra)
library(nycflights13)
library(magrittr) # utils for pipe %>%

##########################

theme_set(theme_bw())

panderOptions('big.mark', ',')
```


# ExercisesS

For this project, we focus on the following datasets: `airlines`, `airports`, `flights`, `planes`, and `weather`.We are going to invstigate these datasets to answeer the following questions:

* Exercises
    + Construct a barplot displaying number of flights per month.
    + Now, in the barplot (showing number of flights per month), make a separate bar for each origin.
    + List of the number of flights for each origin per month
    + What are the top-10 destinations and how many flights were made to these?
    + How many weather observations are there for each origin?
    + Convert temperature to degrees Celsius. This will be used in the reminder of this miniproject.
(We do this for both `temp` and `dewp` using `mutate_at`)

    + Construct a graph displaying the temperature at `JFK`.
    
    + Investigate if arrival delay is associated with the flight distance (and also departure delay).
    
    + Investigate if departure delay is associated with weather conditions at the origin airport. This includes descriptives (mean departure delay), plotting, regression modelling, considering missing values etc.
    
    + Is the age of the plane associated to delay?
    
    
# Solutions 

We start by an exploratory Data to  understand the different datasets so that one can get an idea about how to answer the different questions.

```{r, echo = FALSE}
#library(nycflights13)

airlines
str(airlines)
airports
str(airports)
flights
str(flights)
planes
str(planes)
weather
str(weather)
```
 
  + Construct a barplot displaying number of flights per month.
  
```{r exo_1, echo= TRUE}

# Here, we convert these variables to the categorical ones.

flights$month <- as.factor(flights$month)
flights$origin <- as.factor(flights$origin)
flights$dest <- as.factor(flights$dest)

planes$manufacturer <-as.factor(planes$manufacturer)

# Vector Name of months

v_months <- c("1" = "Jan", "2" = "Feb","3" = "Mar", 
              "4" = "Apr", "5" = "May", "6" = "Jun",
              "7" = "Jul", "8" = "Aug","9" = "sep",                 
              "10" = "Oct","11" = "Nov", "12" = "Dec")

```
Now we can construct a barplot displaying number of flights per month.
```{r echo=TRUE}

nfly_per_month <- flights%>%
  group_by(year, month)%>%
  ggplot(aes(x = month))+
  geom_bar() +
  labs(x = "Months", 
       y = "Number of flights",
       title = "Number of flights per month")+

  scale_x_discrete(labels = v_months)  # Names of months as labels

 # nfly_per_month

 ggplotly(nfly_per_month)  # to makee it interactive.
 
```
  + Now, in the barplot (showing number of flights per month), we can make a separate bar for each origin.

```{r echo =TRUE}
origin_flights <- flights%>%
  ggplot(aes( x = month))+
  geom_bar(aes(fill = origin), position = "dodge") + 
  labs(x = "Months",
       y = "Number of flights", 
       title = "flights for each origin per month") +
  scale_x_discrete(labels = v_months)

 ggplotly(origin_flights) # make it interactive
 
#  origin_flights

```
  + List of the number of flights for each origin per month
```{r echo=TRUE, warning=FALSE}
months_list <- flights%>%
  group_by(month, origin)%>%
  summarise(n_of_flights = n())


 # months_list%>%
  #  pander(big.mark = '.', justify = c('left','center', 'right'))
```

```{r echo= TRUE}
kable(months_list, "html") %>%
  kable_styling("striped","hover") %>%
  scroll_box(width = "500px", height = "600px")
```
  + What are the top-10 destinations and how many flights were made to these?
  
```{r echo= TRUE}
# The top-10 destinations and number of flights to these.

top10 <-flights%>%na.omit()%>%
  group_by(dest)%>%
  summarise(n_flight= n())%>%
  arrange(desc(n_flight))%>%
  select( dest, n_flight)%>%
  head(10)

# List of top-10 destinations and number of flights to these.

top10%>%
  pander(big.mark = '.', decimal.mark = ',', justify = c('left', 'right'))

```
Now, we can order the bars (destinations) according to the total number of flights to these destinations.


```{r echo=TRUE}

nfly_to_top10 <- top10 %>% 
ggplot( aes(x= reorder(dest,-n_flight),n_flight))+
  geom_bar(stat = "identity")+
  labs(x = "Top-10 destinations",
       y = "Number of flights",
       title = "Number of flights")
# nfly_to_top10

ggplotly(nfly_to_top10) # To make it interactive

```
  + How many weather observations are there for each origin?

```{r  echo= TRUE}
# Compute the number of observations for each origin.
weather%>%
  group_by(origin)%>% na.omit()%>%
  summarise(n_observ = n())%>% # Number of observations
  arrange(desc(n_observ))%>%
  pander(big.mark = ',', justify = c('left', 'right'))

```
  + Convert temperature to degrees Celsius. This will be used in the reminder of this miniproject.
(We do this for both `temp` and `dewp` using `mutate_at`)


```{r echo = TRUE}
# Convert degree fahrenheit to degree celcius.
fahrenheit_to_celcius <- function(df)
{
  t_celcius <- (df - 32)* 5/9 
    
    return(t_celcius)
}

# Convert temp and dewp to degree selcius.
weather_dc <-  mutate_at(weather,    vars(temp,dewp),funs(fahrenheit_to_celcius))

head(weather_dc)


# convert variables  origin and month to fator. 

weather_dc$origin <- as.factor(weather_dc$origin)
weather_dc$month <- as.factor(weather_dc$month)

```
  + Construct a graph displaying the temperature at `JFK`.

```{r echo=TRUE}
 jfk_temp <- weather_dc%>%

  group_by(year, month, day)%>%
  filter(origin =="JFK")%>%
  ggplot(aes(month, temp))+
  geom_boxplot() +
  labs(x = "Months",
       y = "Temperature in degree C",
       title = "Graph temperature at JFK") +
  scale_x_discrete(labels = v_months)+
  
# added after submission
  #  geom_boxplot(alpha=0.4) +
    stat_summary(fun.y=mean, geom="point", shape=4,
                 size=2, color="red", fill="red") +
    theme(legend.position="none") 
# end added

 #jfk_temp
  ggplotly(jfk_temp)

```

Now, visualize the daily mean temperature for each origin airport.

```{r echo=TRUE}
mean_temp_orign <- weather_dc%>%
  group_by(origin,year,month, day)%>%
  summarise(mean_temp = mean(temp, na.rm = TRUE))%>%
  ggplot(aes(origin,mean_temp))+
  geom_boxplot(aes(fill = origin))+
  
  # added after submission
  
  #  geom_boxplot(alpha=0.4) +
    stat_summary(fun.y=mean, geom="point", shape=4, size=2, color="red", fill="red") +
  #  theme(legend.position="none") +
# end added
  
  
  
  #geom_point()+
  
  labs(x = "Origins ",
       y = "Mean temperature",
       title = "Mean temperature ")
  
 #mean_temp_orign

ggplotly(mean_temp_orign)

```

  + Investigate if arrival delay is associated with the flight distance (and also departure delay).

In order to investigate if there is any statistical dependence or association between some variables, one can calculate the correlation between these variables. To do so, we will first investigate if these variables are numeric and  if there is no missing values.

```{r echo = TRUE}
 select_cols <- flights%>%
  select(arr_delay, dep_delay, distance)  # Select relevant columns from flights

#Select the rows with  complete cases. 

 flights_complet_cases <- select_cols[complete.cases(select_cols),]

```

Calculate the correlation between  variables
```{r echo = TRUE}
# Correlation test between different variables or columns.

cor_test1 <- cor(flights_complet_cases)

cor_test_tb <- as.data.frame(cor_test1)

 cor_test_tb
 
```
Format the correlation data
```{r echo= TRUE}
   kable(cor_test_tb, "html") %>% # format correlation table.
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "bordered", "responsive"))
```
We can also plot the correlation data to visualize the relation between variables.

```{r echo = TRUE}
corrplot(cor_test1, method="color", type = "upper",
         tl.srt = 45,addCoef.col = "black")

```

 As we can see from the different graphs and correlation parameters , there is 
 a strong correlation between arrival delay and departure delay. In the contrast, there is a negative correlation between the distance and arrival delay. In the next step, I will build a linear regression model using these three variables.
 
 
Now, we are going to build the  Linear model arrival delay and departure delay

```{r echo = TRUE}
# Model arrival delay and departure delay

 del_model1 <- lm(arr_delay ~ dep_delay, data = flights_complet_cases)
coef(del_model1)
summary(del_model1)
```
from this result, one can conclude that departure delay is an important predictor of arrival delay. The changes of departure delay can explain the changes in arrival delay.
```{r echo=TRUE}
pred_delay <- flights_complet_cases%>%
  
  add_predictions(del_model1)%>%
  
  add_residuals(del_model1)
head(pred_delay, 10)
  
  # After submission of report
  pred_plot <- pred_delay%>%
    ggplot(aes(dep_delay, pred))+
    geom_line()+
    geom_point(aes(dep_delay, arr_delay),
               col = "red")  

pred_plot

```

```{r}
# One can also investigate a possibility of quadratisc and polynomial relationship between arrival delay and departure delay and compare the results
my_flights  <- flights%>%na.omit()

  p1_delay =  lm(arr_delay ~dep_delay, data = my_flights)
  
  p2_delay =  lm(arr_delay ~dep_delay
                 + I(dep_delay^2), data = flights)
  
  p3_delay =  lm(arr_delay ~ poly(dep_delay, 5),
                 data = my_flights)
  
 
 p <- qplot(dep_delay, arr_delay, data=my_flights)
  p + geom_line(aes(dep_delay, predict(p1_delay)), col='blue') +
    geom_line(aes(dep_delay, predict(p2_delay)), col='red') +
    geom_line(aes(dep_delay, predict(p3_delay)), col='green') 
  
 
```

  + Investigate if departure delay is associated with weather conditions at the origin airport. This includes descriptives (mean departure delay), plotting, regression modelling, considering missing values etc.

```{r  echo = TRUE}
# We selct the relevant columns from  both flights and weather datasets
# then join  the two subsets.

flight_delay <- flights%>%
  select(year:day, origin, hour, arr_delay, dep_delay, tailnum)
#flight_delay
```
Join subset flights and planes datasets
```{r echo= TRUE}
fly_w_cd <-  inner_join(flight_delay, weather_dc,
                        by = c("origin" = "origin",
                                 "year" = "year",
                                "month" = "month",
                                  "day" = "day",
                                 "hour" = "hour"))

#c("year", "day","month", "hour", "origin")

# Here  I consider only the completecases from dataset.

 fly_weather_cd <- fly_w_cd[complete.cases(fly_w_cd),]
 
head(fly_weather_cd)

```
Now we can compute the correlation between numeric relevant variables.

```{r echo = TRUE}
num_cols <- sapply(fly_weather_cd, is.numeric)

fly_weather_cd  <- fly_weather_cd[, num_cols] # numeric columns

delay_weather_cd <- fly_weather_cd%>%
  select(dep_delay, humid,precip, visib,
         wind_dir, wind_speed, temp, dewp)

cor_delay_wc <- cor(delay_weather_cd)

cor_delay_wc
```
 We can Plot  the correlation data to see the relations between variables
 
```{r echo = TRUE}
corrplot(cor_delay_wc, method="color", type = "upper",
         tl.srt = 45,addCoef.col = "black")

#corrplot(cor_delay_wc, method = 'number')
```
 
According to correlation plots, there is no strong correlation between departure delay and any of the weather conditions variables and there is a tight competition between positive and negative correlations. To get more information we will visualize the relationship between delay and some of the variable such as `temp` and ``precip`. 
 
```{r echo = TRUE}
fly_w_cd%>%
  group_by(precip)%>%
  summarise (dp_delay = mean(dep_delay, na.rm = TRUE))%>%
  ggplot(aes(precip, dp_delay))+
  geom_line()+
 
  geom_point()

```
 As we can see from the graph, the precip doesn't provide more information explanation about departure delay and the correlation between these two variables is weak.
```{r echo = TRUE}
fly_w_cd%>%
  group_by(temp)%>%na.omit()%>%
  summarise (dp_delay = mean(dep_delay, na.rm = TRUE))%>%
  ggplot(aes(temp, dp_delay))+
 # geom_line(col = "red")+
 
  geom_point()+
 geom_smooth(  se = FALSE, method = 'loess')

```
Surprisingly, can that the departure delay increase with temperature but this variation is not significant to provide more explanation about departure delay changes. In the next step I will build the linear regression model.

Modelling weather conditions delay model

```{r echo = TRUE}
set.seed(101)
#delay_weather_cd%>%
wc_delay_sample <- sample.split(delay_weather_cd$dep_delay, 
                                SplitRatio = 0.7)

# train set
train <- subset(delay_weather_cd, wc_delay_sample == TRUE)
head(train)
#Test set
test <- subset(delay_weather_cd, wc_delay_sample == FALSE)
head(test)
```
 Build linear model 
```{r echo = TRUE}
model <- lm(dep_delay~ ., data = train)
summary(model)

# Residuals

residus <- residuals(model)
residus <- as.data.frame(residus)
#head(residus)

```

```{r}
ggplot(residus, aes(residus))+
geom_freqpoly(binwidth = 0.5)

```


```{r echo = TRUE}
residus%>%
ggplot(aes(residus)) +  geom_histogram(fill='blue',alpha=0.5, bins = 20)
```


```{r echo = TRUE}
# After submission of report

delay_weather_cd%>%
ggplot(aes(dep_delay)) +  geom_histogram(fill='blue',alpha=0.5, bins = 20)
```


Plotting model
```{r}
#plot(model)
```
Prediction computation


```{r echo = TRUE}
delay_predict <- predict(model, test)
predict_res  <- cbind(delay_predict, test$dep_delay)

colnames(predict_res) <- c("predict", "actual")

predict_res<- as.data.frame(predict_res)

 head(predict_res)
 
 head(delay_predict)
 
```

 + Is the age of the plane associated to delay?

Here, we start by computing the age of each and store it in new column names ` age`.
```{r echo = TRUE}

# This function compute the age the plane as integer.
cur_year <- as.integer(format(Sys.Date(), "%Y"))

year_diff <- function(x_year, y_year = cur_year)
{
   x_year <- as.integer(x_year)
  
    
 return(y_year - x_year)
 
}
```



```{r echo = TRUE}
pl_ages <- planes%>%
 mutate(age = year_diff(year))%>%
  select(tailnum, age, seats)
pl_ages <- pl_ages[complete.cases(pl_ages),]


flight_delay <- flights%>%
  select(year:day, origin, hour, arr_delay, dep_delay, tailnum)

#  join planes and flights datasets

join_fly_planes <- flight_delay%>%
  inner_join(pl_ages, by = "tailnum")
  #join_fly_planes
  
 fly_planes <-join_fly_planes%>%
   select(age,dep_delay,arr_delay)
 fly_planes <-fly_planes[complete.cases(fly_planes),]
 fly_planes

```
 
 Now I will perform  some plotting to  visualize the graphic (age vs delay)

```{r echo = TRUE}
age_delay <- fly_planes%>%
  group_by(age)%>%
  filter(!is.na(dep_delay)) %>%
  summarise(delay = mean(dep_delay)) %>%
  ggplot(aes(x = age, y = delay)) +
 # geom_point() +
  geom_line() +
  labs(x = "Age of planes",
       y = "Mean of departure delay",
       title = "Graph dep_delay vs age")
 age_delay
#ggplotly(age_delay)

```
try correlation between  age of planes and arrival delay.
```{r echo = TRUE}

cor_delay_age  <-cor(fly_planes)

cor_delay_age

```
According to the correlation table, There is a weak negative correlation between age and both arrival delay and departure delay. Meaning, the age of the planes can not explained the changes of delay.  

Format the correlation data

```{r}
   kable( cor_delay_age, "html") %>% # format correlation table.
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

  + It seems like the plane manufacturer could use a cleaning. After that, how many manufacturers have more than 200 planes? And how many flights are each manufacturer with more than 200 planes responsible for?

```{r echo = TRUE}
# Find list of manufacturers that have more than 200 planes.
manufact_200_planes <- planes%>%
  filter(!is.na(manufacturer)) %>%
group_by( manufacturer)%>%
  summarise(n_planes =n())%>%
  arrange(desc(n_planes))%>%
  filter(n_planes > 200)
#print list 
  manufact_200_planes%>%
    pander(big.mark = ',', justify = c('left', 'right'))
  
 # Combine the subset manufact_200_planes with flights dataset to find number of flights.
  
num_planes200 <- planes%>%
  inner_join(flights,"tailnum")%>% 
  filter(manufacturer %in%
           manufact_200_planes$manufacturer)%>% 
  
  group_by( manufacturer)%>%
  summarise(n_flight =n())%>%
  arrange(desc(n_flight))

  # print list
  num_planes200%>%
     pander(big.mark = ',', justify = c('left', 'right'))

 # format number of flights
  
 kable(num_planes200, "html") %>% # format correlation table.
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", "bordered"))

```

It turns out that Airbus has several main models, e.g. several A320 (A320-211, A320-212 etc.) and so on. Create a frequency table of the main models for Airbus and how many planes there are in each.

```{r echo = TRUE}
# Find number of model for "AIRBUS"

airbus_models <- planes%>% 
  group_by(model)%>%
  filter(manufacturer =="AIRBUS")%>%
  summarise(n_of_planes = n())%>%
  arrange(desc(n_of_planes))

# Split column model to find main models
#airbus_models%>%
  
  split_m <- strsplit( airbus_models$model, "-")

main_mdl <- vector("character", length = length(split_m))

for (m in 1: length(split_m)) {
  
  main_mdl[m] <- split_m[[m]][1]
  
}
# List of main models

m_models <- airbus_models%>%
  mutate(main_model =  main_mdl)

 m_models %>%
  pander(big.mark = ',', justify = c('left','center', 'right'))

 # format list
  
 kable(m_models, "html") %>% # format correlation table.
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", "bordered"))
  

```


```{r echo = TRUE }
# A Frequency table of main models

m_models%>%
  group_by(main_model)%>%
  summarise( num_model = n(), num_of_planes = sum(n_of_planes))%>%
pander(big.mark = ',', justify = c('left','center', 'right'))
```


## Exercise

Are larger planes (measured by number of seats) more or less delayed than smaller planes?
To answer this question one can visualize the plot (` seats vs arr_delay`)

```{r echo = TRUE}
delay_seats <- join_fly_planes%>%
  group_by(seats)%>%
  summarise(mean_delay = mean(arr_delay, na.rm = TRUE))%>%
  arrange(desc(seats))

delay_seats%>%
pander(big.mark = ',', justify = c('left', 'right'))

kable(delay_seats, "html") %>%
  kable_styling() %>%
  scroll_box(width = "500px", height = "600px")

```

```{r}

delay_seat<- delay_seats%>%
  ggplot(aes(factor(seats), mean_delay))+
  geom_point() +
  labs(x = "Number of seats",
       y = "Mean of departure delay",
       title = "Graph dep_delay vs age")
 
delay_seat
#ggplotly(delay_seat)
```

According to this graph, there is no sign that the larger planes are more delayed than the smaller ones.

One can also try to compute the correlation between ` age, arr_delay` and ` dep_delay`.

```{r}
seat_delay <- join_fly_planes%>%
  select(seats, arr_delay, dep_delay)%>% na.omit()

 seat_cor <- cor(seat_delay) # correlation
 
corrplot(seat_cor, method="color", type = "upper",
         tl.srt = 45,addCoef.col = "black")
   
   kable(seat_cor, "html") %>% # format correlation table.
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive", "bordered"))
   
```
 There is a weak negative correlation between a number of seats and delay meaning that the number of seats (smaller/ larger) can not explain the arrival delay.
 
```{r}
 set_delay_model <- lm(seat_delay$arr_delay ~seat_delay$seats)
summary(set_delay_model)
coef(set_delay_model)
```

## Exercise

On a map (`map_data("usa")`), plot the airports that has flights to them.

```{r echo=TRUE}
usa <- map_data("usa")


#airports$faa <- as.factor(airports$faa)

fli_dest <- flights%>%
  group_by(dest)%>%
  summarise(n = n())%>%
  inner_join(airports, by =c(dest = "faa"))
fli_dest 


```

```{r}

ggplot() + 
  geom_polygon(data = usa, aes(long, lat, group = group), fill = "grey") + 
  geom_point(data = fli_dest , aes(lon, lat), color="blue") + 
  labs(x = "longitude",
       y = "latitude",
       title = "map flights for destinations")+ 
  borders('state')+
  coord_quickmap()

```

Make a similar plot, but now points must have size relative to the number of flights each airport is destination for. 


```{r echo = TRUE}

map_data <- ggplot() + 
  geom_polygon(data = usa, aes(long, lat, group = group), fill = "grey") +

geom_point(data =fli_dest, aes(x = lon, y = lat, size = n , color = n)) + 
  borders("state")+
  scale_color_viridis()+
  
  labs(x = "longitude",
       y = "latitude",
       title = "map flights for destinations")+
coord_quickmap()
map_data
#ggplotly(map_data)

```

## Exercise

Do a principal component analysis of the weather at JFK using the following columns: `temp, dewp, humid, wind_dir, wind_speed, precip, visib` (only on `complete.cases()`). How many principal components should be used to capture the variability in the weather data? 

```{r echo=TRUE}
weather_dc <-  mutate_at(weather, vars(temp, dewp),funs(fahrenheit_to_celcius))

weather_jfk_df <- weather_dc%>%
  group_by(origin)%>%
filter(origin == "JFK") %>%
   select(origin,temp, dewp, humid, wind_dir, wind_speed, precip, visib)

head(weather_jfk_df)

```

```{r echo = TRUE}
# Here we consider the completecases

weather_jfk<- weather_jfk_df[complete.cases(weather_jfk_df),]
 
weather_jfk <-  weather_jfk[, sapply(weather_jfk, is.numeric)]

```

summary and correlation plot

```{r}
cor_data <- cor(weather_jfk)
corrplot(cor_data, method="color", type = "upper",
         tl.srt = 45,addCoef.col = "black")
```

Principal component analysis PCA

```{r}
pairs(weather_jfk)
ggpairs(weather_jfk)

head(weather_jfk)
summary(weather_jfk)

```


```{r}
weather_jfk_pca <- prcomp(weather_jfk, scale. =  TRUE)

fviz_eig(weather_jfk_pca)

#plot(weather_jfk_pca)
plot(weather_jfk_pca, type = 'l')
summary(weather_jfk_pca)
weather_jfk_pca

```
 
```{r cache= TRUE}

ggpairs(weather_jfk_pca$x %>% as_tibble())
```


Accordingly, to the different plots and summary table, one can see that more than 94% of weather variability can be captured by 5 PCs. So we choose the 5 first components as principal ones. 

```{r, fig.width = 12, fig.asp=1}
par(mfrow = c(3,3), mar = c(1,1,1,1)*2)

biplot(weather_jfk_pca, scale = 0, choices = c(1,2))
#biplot(weather_jfk_pca, scale = 0, choices = c(1,3))
#biplot(weather_jfk_pca, scale = 0, choices = c(2,3))
#biplot(weather_jfk_pca, scale = 0, choices = c(2,4))
#biplot(weather_jfk_pca, scale = 0, choices = c(3,4))
#biplot(weather_jfk_pca, scale = 0, choices = c(4,5))
#biplot(weather_jfk_pca, scale = 0, choices = c(5,3))

```



  